Technologies of Semantic WEB as an environment of application development and integration

### 2022-02-04 paper-aicts-cherkashin.tex
Grammarly check done

%%
%% One can fix some overfulls
\sloppy

@@

Since 2001 \cite{tbl}, the Semantic Web (SW) technologies have become actively used in describing software domains, representing semantic relations between objects. For some domains on the base of the SW description, first-order knowledge theories are constructed, but in the practical aspect of the SW utilization, the solving data representation and software integration aspects prevail. Some tools for automation of conceptual model design and model conversion were proposed \cite{hogan}. The R\&D at present are carried on in developing applications, improving and standardizing vocabularies (conceptual domain models), devising tools for solving particular problems.

@@

Namespaces are convenient instruments in the triple definition. For example \verb|integer| literal \verb|42^^<http://www.w3.org/2001/XMLSchema#integer>| can be reduced to \verb|42^^xs:integer|, where \verb|xs| is namespace prefix of term \verb|integer|.  Most KG representation formats allow one to define a default namespace, which is used without prefix. Almost any conceptual model of a domain represented as KG comprises entities taken from multiple vocabularies, and namespaces specify the origin of a term. Three-component tuple (triple) outside its KG context sometimes added the fourth KG IRI component (quadruple).

@@

KGs and their subgraphs are programmatically processed using libraries developed for all papular programming environments. KG is represented with a triple or quadruple set object. The set is traversed, triples are filtered according to various conditions. Filters are usually defined as patterns, where each component is either a free variable/placeholder or a concrete value (IRI, literal), which must match exactly.   Some libraries allow one to execute \texttt{SPARQL} queries locally. KG is evolved by adding new triples and removing irrelevant ones. The resulting modified KG is stored as files and published on a server. An example of a federated SPARQL query is presented in Figure~\ref{fig:sparql-ex1}, which results in counting parts classes as the first level of a three widget. Each class is named in Russian if \texttt{DBPedia} contains a corresponding label with \texttt{ru} language tag.  Otherwise, the label is figured out of IRI. This query is initially run at a local server, and the server initiates a subquery to \texttt{DBPedia.org} endpoint.

@@

Authoring tool (B1) is used to provide document generation and integration \cite{zont19}. This tool enables document content and structure inference in the browser utilizing templates and data from other web pages and websites providing LOD\footnote{Abbreviation of Linked Open Data} \cite{lod} capabilities. Programmer implements a view of an object as a web page with an extended version of RDFa. The page is generated within any programming runtime, \emph{e.g.} PHP. The authoring tool engine collects web page chunks and composes the document as an XHTML page. This composed page can be stored in a KG as a set of triples and indexed in the Text indexing module (C1), printed with the browser's printing engine, or converted to PDF. The tool allows the user to change document content in WYSIWYG mode. Its general structure is defined by changing the template on the server-side.

@@

In this section, three examples of application implementation are being presented.  The first example is an MDA application in the synthesis of a Next-generation sequencing (NGS) investigation environment.  After the invention of methods of NGS and their introduction in practice of biological systems research, a new direction of molecular genetics is formed, which is referred to as \emph{metagenomics}.  The method allows biologists to describe a significant number of new groups of organisms at all taxonomic levels.

@@

The aim of this R\&D \cite{zont19, aicts2020} is to develop software support of the processes of NGS analysis with techniques and software for a visual representation of the computational process of so-called amplicon analysis so that domain specialists would be able to compose computational pipelines, which are executed on distributed heterogeneous computing resources (clouds).  The visual models, later, are to be used for organizing the computational process and data store and articular retrieval services.

The tools are based on dataflow representation of various stages of NGS standard operation procedures (SOPs)  formed out of individual operations of Mothur library in Rapidminer studio Java application.  To reflect Mothur's operations we implemented source code analysis and transformation software.  The C++ sources were analyzed by Python syntax analyzer based on regular expressions and a top-level algorithm, generating a \verb|TTL| source of KG representing Mothur modules' interfaces.  That was a Computation-independent model (CIM) in the implemented MDA.  CIM transformed into the Platform-independent model (PIM), represented Mothur's modules as dataflow modules, in particular, differencing parameters to input/output dataflow ports and parameters.  The Platform-specific model (PSM), \emph{i.e.}, models of Java sources of the dataflow blocks, was constructed, in turn, from PSM.

All transformations was carried out with object-oriented (OO) Logtalk represented set of modules that logically infer new data descriptions of the target objects.  Thanks to OO, the transformation have two level of abstraction in its representation.  The higher one is a network of objects sending messages to each other via their interfaces.  The lower is the set of rules analyzing and producing new conclusions on generated object properties.

The PSM objects must have an order in their description as they represent a models of source code to be generated.  The order is organized with blocks of \verb|assert|ed statement and subobjects sequences referencing KG elements.  Blocks can be appended and pretended with new elements.  At the stage of rendering sources, the three of blocks are traversed in-depth.  This leverages known \verb|llvmlite| open-source approach of code generation.  Figure~\ref{fig:block-seq} shows the main idea behind PSM object synthesis.
\begin{figure}
  \centering
\begin{multicols}{2}
\begin{minted}[fontsize=\scriptsize]{logtalk}
:- object(class, specializes(code_block),
   imports([named])). % Category of named entities
:- public([classlist/1, methods/1, attributes/1]).
% . . . . . . . . . . . . . .
renderitem(Item, Result):-      % proceed with default
    ^^renderitem(Item, Result). % rendering
render(Result):-         % Source generator
    ^^render(Name),      % implemented in a category
    ( ::item(classlist(List)) ->
     % . . . . . . . . . . .
        [Name]) ),
    ( ::item(attributes(Attributes))->
     % . . . . . . . . . . .
        [DefAttrList]),
      Attributes::items(InstanceAttrs),
      findall(S, ( % initialize attributes
         % . . . . . . . . .
         ), AttrAssigns),
        root::unindent,
        AttrList=[ConstructorDef|AttrAssigns];
         % . . . . . . . . .
        AttrList=[ConstructorDef, Pass] ),
    ( ::item(methods(Methods))-> % If any ...
      Methods::render(MethodList);
      MethodList=[] ),
    lists::append(AttrList,MethodList,StringList),
    root::unindent, Result=[Signature|StringList].
:- end_object.
\end{minted}
      \includegraphics[width=0.7\linewidth]{code_block_class.pdf}
\end{multicols}
  \caption{Procedure of dataflow module source code model synthesis}
  \label{fig:block-seq}
\end{figure}
This example of the application of KG utilizes strongly the property of data representation flexibility for description of the software object model data as well as one central point of data store and access.  The disadvantageous properties of RDF, namely, the weak expressiveness with triples, was neglected by predicate and object encapsulation.

@@

This project \cite{authoring} is aimed at construction of an environment for document authoring out of user editable template elements, database query results, LOD-compliant web-pages, and other document data.  Target document is a web page which content is inferred while loaded by the browser.

All document elements are either a context structures of templates filled with other elements.  A context is a RDF hierarchical representation of a document data structure.  It provides data to be filled in a current rendered template.  Document is a main template.  All elements are SW resources.  Templates can contain JavaScript subroutines for implementing special service, for example, table of contents out of CSS \verb|class| labeled subsections of the generated document.

All elements are stored either in a KG as triples, or as LOD-compliant web-pages, and according to SW consideration all of them form a distributed KG.  Data access differentiation by security model is implemented programmatically as the web pages disallowing access to the restricted information.  SW has special predicates for description a possibility and location of a private data sub-KG in the open part.

@@

The general logical structure (sections, subsections, \emph{etc.}) are represented with Open Annotation (\textbf{oa}) vocabulary originally used for annotation of already existing content. Using \textbf{oa}, the generated document is converted to PDF and other formats.  HTML5 default tag structures are used for visual sectioning and style definition.

The \textbf{oa} vocabulary also used to define parts of document, which can be interactively edited after the document have been already rendered.  The obtained changes are fixed in a commit as new resources, and the containing document obtains new IRI fixing all the set of resource changes.

@@

Project \cite{gisviewer} devoted to investigate capabilities of SW and KG technologies in representation spatially distributed data (GIS\footnote{Abbreviation of Geographical Information System}-data).  The fault data \cite{lunina,afs} are chosen as subject for representation and publishing in this investigation.  In the geology, researchers accumulate data obtained after event observations, \emph{e.g.}, earthquakes, landslides, by analyzing remote sensing data and results of field works.  The obtained data are processed and interpreted, resulting in setting new attributes to a fault or refinement of their values.  According to the techniques of geological research, additional information are associated with attributes, clarifying their values.  Such clarification comprises so-called precision characteristics, measurement conditions, reliability assessment, and paper references, where fault data were published.

GIS represents spatial data in semantic-defined layers.  For each object of a GIS layer, one can associate a set of attribute values of auxiliary data.  The set of the attributes are the same for each object of the layer, regardless of whether the attribute value is defined for an object or not.  Empty values are represented as ``\texttt{null}''s.  In the case of geological exploration, when a lot of attributes are undefined, this approach leads to sparse filled tables.  This, in turn, requires data modification and analysis algorithms to utilize additional data checking stages when using standard relation operations (SELECT, UPDATE, DELETE).

Web publication applications, as information systems, are to implement filtering functions, differentiating the value attributes and their metadata.  Screen widgets label names either defined in application configuration or figured out from the attribute names.  Lack of vocabulary formal domain definition forces developer to spend more efforts for the user interface implementation.

In this project, we constructed a Web-application using contemporary Web and SW technologies for rendering a fault map, which visual content is controlled by data filters implemented as SPARQL queries to KG representing fault data.  For the purpose, a data conversion from the tabular form to KG was carried out, a CliopPatria server for data storage was installed, and data viewers for the triple sets for a common subject (a fault) has been implemented.  In Figure~\ref{fig:gis-ex} an example of the filtering and a selected object data viewing form is presented.
\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{faults-leaflet-doc.png}
  % \caption{Viewer window showing description of a fault}
  \caption{Example of GIS rendering of KG data}
  \label{fig:gis-ex}
\end{figure}

The proposed technology and software allows one to construct Web-GIS systems for research communities, as they support constant data accumulation, aggregation and analysis thanks to the properties of knowledge graph (KG) data storage and processing.

@@

Standardized forms of basic concepts of data representation and processing in the filed of Sematic Web (SW) research were considered in this paper.  A special attention has been paid to reviewing Knowledge Graph (KG) data storage and processing in scientific data processing software.

We did not consider analogous R\&D as it was already reviewed in the corresponding our publications.  Another point of our R\&D is a strategy for improving existing technologies, data and software with our tools, such as MDA.  We present a review of obtained SW and KG technologies application experience.  The main one is the powerful capabilities of application integration, where all the set of implemented software, in fact, form a distributed environment of data processing.  The main requirement is that the data representation at storage services and in the protocol implementations should obey SW and KG standards.

The main advantages of these technologies for the software developer are semantic markup and generic properties of KG, especially a possibility to postpone formal data structure definition as such declarative data are easily can be processed with a rule-based inference engine.  Scientific community results in domain description in standardized vocabularies (ontologies), user communities aggregating and formalizing data on real-world objects form a productive background for problem space description and,  in its turn, a common basis for application integration, which result in a cumulative effect of aggregation of developed software in a complex distribute data processing system and environments.

Further development of the technologies include the following directions of a great interest:
\begin{itemize}
\item Fibered representation of KG to support consistency and completeness check by reflecting A-Box entities and their relations to template domain models;
\item Investigate expressive capabilities of Logtalk programming language as an instrument of knowledge (rules) representation for KG processing modules;
\item Adding extended document analysis services \cite{shigpaper} to the environment (section~\ref{sec:architecture}), which are developed in our research group;
\item Improve integration techniques with metamodel presentation of the execution environment.
\end{itemize}

>>>>

%%
%% One can fix some overfills
\sloppy

@@

Since 2001 \cite{tbl}, the Semantic Web (SW) technologies have become actively used in describing software domains, representing semantic relations between objects.  First-order knowledge theories are constructed for some domains based on the SW description, but in the practical aspect of the SW utilization, the solving data representation problems and software integration aspects prevail.  Some tools for automation of conceptual model design and model conversion were proposed \cite{hogan}.  The R\&D at present are carried on in developing applications, improving and standardizing vocabularies (conceptual domain models), devising tools for solving particular problems.

@@

Namespaces are convenient instruments in the triple definition.  For example \verb|integer| literal \verb|42^^<http://www.w3.org/2001/XMLSchema#integer>| can be reduced to \verb|42^^xs:integer|, where \verb|xs| is namespace prefix of term \verb|integer|.  Most KG representation formats allow one to define a default namespace used without prefix.  Almost any conceptual model of a domain represented as KG comprises entities taken from multiple vocabularies, and namespaces specify the origin of a term.  Three-component tuple (triple) outside its KG context sometimes added the fourth KG IRI component (quadruple).

@@

KGs and subgraphs are programmatically processed using libraries developed for all papular programming environments.  KG is represented with a triple or quadruple set object.  The set is traversed, triples are filtered according to various conditions.  Filters are usually defined as patterns, where each component is either a free variable/placeholder or a concrete value (IRI, literal), which must match exactly.   Some libraries allow one to execute \texttt{SPARQL} queries locally.  KG is evolved by adding new triples and removing irrelevant ones.  The resulting modified KG is stored as files and published on a server.  An example of a federated SPARQL query is presented in Figure~\ref{fig:sparql-ex1}, which results in counting parts classes as the first level of a three widget.  Each class is named in Russian if \texttt{DBPedia} contains a corresponding label with \texttt{ru} language tag.  Otherwise, the label is figured out of IRI.  This query is initially run at a local server, and the server initiates a subquery to \texttt{DBPedia.org} endpoint.

@@

Authoring tool (B1) is used to provide document generation, and integration \cite{zont19}.  This tool enables document content and structure inference in the browser utilizing templates and data from other web pages and websites providing LOD\footnote{Abbreviation of Linked Open Data} \cite{lod} capabilities.  Programmer implements a view of an object as a web page with an extended version of RDFa.  The page is generated within any programming runtime, \emph{e.g.} PHP.  The authoring tool engine collects web page chunks and composes the document as an XHTML page.  This composed page can be stored in a KG as a set of triples and indexed in the Text indexing module (C1), printed with the browser's printing engine, or converted to PDF.  The tool allows the user to change document content in WYSIWYG mode.  Its general structure is defined by changing the template on the server-side.

@@

In this section, three examples of application implementation are being presented.  The first example is an MDA application in synthesizing a Next-generation sequencing (NGS) investigation environment.  After the invention of methods of NGS and their introduction in the practice of biological systems research, a new direction of molecular genetics is formed, which is referred to as \emph{metagenomics}.  The method allows biologists to describe many new groups of organisms at all taxonomic levels.

@@

The aim of this R\&D \cite{zont19, aicts2020} is to develop software support of the processes of NGS analysis with techniques and software for a visual representation of the computational process for so-called amplicon analysis so that domain specialists would be able to compose computational pipelines, which are executed on distributed heterogeneous computing resources (clouds).  Later, the visual models are used to organize the computational process, data store, and particular NGS data retrieval services.

The tools are based on dataflow representation of various stages of NGS standard operation procedures (SOPs)  formed out of individual operations of Mothur library in Rapidminer studio Java application.  To reflect Mothur's operations, we implemented source code analysis and transformation software.  The C++ sources were analyzed by a Python syntax analyzer based on regular expressions and a top-level algorithm, generating a \verb|TTL| source of KG representing Mothur modules' interfaces.  That was a Computation-independent model (CIM) in the implemented MDA.  CIM transformed into the Platform-independent model (PIM), represented Mothur's modules as dataflow modules, mainly differencing parameters to input/output dataflow ports and parameters.  The Platform-specific model (PSM), \emph{i.e.}, models of Java sources of the dataflow blocks, was constructed, in turn, from PSM.

All transformations were carried out with an object-oriented (OO) Logtalk programming environment representing modules that logically infer new data descriptions of the target objects.  Thanks to OO, the transformation has two levels of abstraction in its representation.  The higher one is a network of objects sending messages via their interfaces.  At the lower one, rules analyze and produce new conclusions on generated object properties.

The PSM objects must have an order in their description as they represent a model of source code to be generated.  The order is organized with blocks of \verb|assert|ed statement and subobjects sequences referencing KG elements.  Blocks can be appended and pretended with new elements.  At the rendering sources stage, the three blocks are traversed in-depth.  It leverages the known \verb|llvmlite| source code generation approach.  Figure~\ref{fig:block-seq} shows the main idea behind PSM object synthesis.
\begin{figure}
  \centering
\begin{multicols}{2}
\begin{minted}[fontsize=\scriptsize]{logtalk}
:- object(class, specializes(code_block),
   imports([named])). % Category of named entities
:- public([classlist/1, methods/1, attributes/1]).
% . . . . . . . . . . . . . .
renderitem(Item, Result):-      % proceed with default
    ^^renderitem(Item, Result). % rendering
render(Result):-         % Source generator
    ^^render(Name),      % implemented in a category
    ( ::item(classlist(List)) ->
     % . . . . . . . . . . .
        [Name]) ),
    ( ::item(attributes(Attributes))->
     % . . . . . . . . . . .
        [DefAttrList]),
      Attributes::items(InstanceAttrs),
      findall(S, ( % initialize attributes
         % . . . . . . . . .
         ), AttrAssigns),
        root::unindent,
        AttrList=[ConstructorDef|AttrAssigns];
         % . . . . . . . . .
        AttrList=[ConstructorDef, Pass] ),
    ( ::item(methods(Methods))-> % If any ...
      Methods::render(MethodList);
      MethodList=[] ),
    lists::append(AttrList,MethodList,StringList),
    root::unindent, Result=[Signature|StringList].
:- end_object.
\end{minted}
      \includegraphics[width=0.7\linewidth]{code_block_class.pdf}
\end{multicols}
  \caption{Procedure of dataflow module source code model synthesis}
  \label{fig:block-seq}
\end{figure}
This example of the application of KG utilizes strongly the property of data representation flexibility for description of the software object model data as well as one central point of data store and access.  The disadvantageous properties of RDF, namely, the weak expressiveness with triples, was neglected by predicate and object encapsulation.

@@

This project \cite{authoring} is aimed at the construction of an environment for document authoring out of user-editable template elements, database query results, LOD-compliant web pages, and other document data.  The target document is a web page where content is inferred while loaded by the browser.

All document elements are either context structures or templates filled with other elements.  A context is an RDF hierarchical representation of a document data structure, providing data to be filled in a current rendered template.  The document is rendered from its main template.  All elements are SW resources.  Templates can contain JavaScript subroutines for implementing special services, for example, table of contents out of CSS \verb|class| labeled subsections of the generated document.

All elements are stored in a KG as triples and as LOD-compliant web pages, and according to SW considerations, all of them form a distributed KG.  Data access differentiation by security model is implemented programmatically as the web pages disallow access to the restricted information.  SW has special predicates for describing a possibility and location of a private data sub-KG in the open part.

@@

The general logical structure (sections, subsections, \emph{etc.}) are represented with Open Annotation (\textbf{oa}) vocabulary originally used for annotation of already existing content. The generated document conversion into PDF and other formats is carried out under the control of \textbf{oa}.  HTML5 default tag structures are used for visual sectioning and style definition.

The \textbf{oa} vocabulary is also used to define parts of a document, which allow interactive editing after the document has been rendered.  The obtained changes are fixed in a commit as new resources, and the containing document obtains new IRI fixing all the set of resource changes.

@@

Project \cite{gisviewer} devoted to investigating capabilities of SW and KG technologies in representation spatially distributed data (GIS\footnote{Abbreviation of Geographical Information System}-data).  The fault data \cite{lunina,afs} are chosen as a subject for representation and publishing in this investigation.  In geology, researchers accumulate data obtained after event observations, \emph{e.g.}, earthquakes, landslides, by analyzing remote sensing data and results of field works.  The obtained data are processed and interpreted, setting new attributes to a fault or refinement of their values.  According to the geological research techniques, additional information is associated with attributes, clarifying their values.  Such clarification comprises precision characteristics, measurement conditions, reliability assessment, paper references, and published fault data.

GIS represents spatial data in semantic-defined layers.  For each object of a GIS layer, one can associate a set of attribute values of auxiliary data.  The attribute set is the same for each object of a layer, regardless the attribute's value is defined or not.  Empty values are represented as ``\texttt{null}''s.  In the case of geological exploration, when many attributes are undefined, this approach leads to sparsely filled tables.  In this case, structure modification and data analysis algorithms are required to utilize additional data checking stages when using standard relation operations (SELECT, UPDATE, DELETE).

Web publication applications, like information systems, implement filtering functions, differentiating the value attributes and their metadata.  Screen widgets label names either defined in application configuration or figured out from the attribute names.  Lack of vocabulary formal domain definition forces developers to spend more effort on user interface implementation.

In this project, we constructed a Web-application using contemporary Web and SW technologies for rendering a fault map, in which visual content is controlled by data filters implemented as SPARQL queries to KG representing fault data.  For this purpose, data conversion from the tabular form to KG was carried out, a CliopPatria server for data storage was installed, and data viewers for the triple sets for a typical subject (a fault) have been implemented.  In Figure~\ref{fig:gis-ex} an example of the filtering and a selected object data viewing form is presented.
\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{faults-leaflet-doc.png}
  % \caption{Viewer window showing description of a fault}
  \caption{Example of GIS rendering of KG data}
  \label{fig:gis-ex}
\end{figure}

The proposed technology and software allow one to construct Web-GIS systems for research communities, as they support constant data accumulation, aggregation, and analysis thanks to the properties of knowledge graph (KG) data storage and processing.

@@

Standardized forms of basic concepts of data representation and processing in Semantic Web (SW) research were considered in this paper.  Special attention has been paid to reviewing Knowledge Graph (KG) data storage and processing in the scientific data processing software.

We did not consider analogous R\&D as it was already reviewed in the corresponding publications.  Another point of our R\&D is improving existing technologies, data, and software with our tools, such as MDA.  We present a review of obtained SW and KG technologies application experience.  The main one is the powerful application integration capabilities, where all the set of implemented software forms a distributed environment of data processing.  The main requirement is that the data representation at storage services and in the protocol implementations should obey SW and KG standards.

The main advantages of these technologies for the software developer are semantic markup and generic properties of KG, especially a possibility to postpone formal data structure definition as such declarative set-based data can be processed with a rule-based inference engine.  Scientific community results in domain description in standardized vocabularies (ontologies), user communities aggregating and formalizing data on real-world objects form a productive background for problem space description and,  in its turn, a common basis for application integration, which results in a cumulative effect of aggregation of developed software in a complex distributed data processing system and environments.

Further development of the technologies includes the following directions of great interest:
\begin{itemize}
\item Fibered representation of KG to support consistency and completeness check by reflecting A-Box entities and their relations to template domain models;
\item Investigate expressive capabilities of Logtalk programming language as an instrument of knowledge (rules) representation for KG processing modules;
\item Adding extended document analysis services \cite{shigpaper} to the environment (section~\ref{sec:architecture}), which are developed in our research group;
\item Improve integration techniques with metamodel presentation of the execution environment.
\end{itemize}
